Conflict Driven Evolution and Urban Development: \\ Evidence from the Holy Roman Empire\footnote{ I thank Lara Andermann, Leonie Bielefeld, Daniel Bursian, Davide Cantoni, Emilio Esguerra, Filip Milojevic, Johannes Vorbach and Matthias Weigand for helpful comments and feedback. }

### 2023-06-27 paper/paper.tex
#32 three-author citations

By contrast, my story is not about population dynamics but institutions (although it is compatible with Malthusianism). A longstanding tradition in economics emphasises the role of institutional factors in long-run growth (e.g. \cite{north1970}, \cite{delong1993}, \cite{ajr2001}). From this vast literature one theme may be especially relevant here: the institution of fiscal capacity, or the state's ability to collect taxes. \cite{tilly1985} argues that the main reason for states to collect taxes is to be able to wage war. In other words, fiscal capacity and state power are tightly linked --- so if conflict selects for state power, then it also selects for fiscal capacity. \cite{gennaioli2015} and \cite{cantoni2023} show that conflict drove state consolidation, with the latter set of authors using the same data as this paper. \cite{dincecco2012} show that fiscal capacity had a positive effect on growth.

@@

The closest relative of this paper is \cite{schoenholzer2022} who investigate how switching between states affects the development of cities. They find that switching states affects population negatively in the short run but positively in the long run. Further, they show that this long-run benefit is driven by improved state quality after switching. Similar to the evolution metaphor, they interpret this as "creative destruction": the short run cost of destruction is compensated by the long-run benefit of becoming part of a higher-quality state. However, they do not show explicitly how these improvements in state quality are driven by selection through conflict. Their data has a bigger geographical scope (all of Europe) but lower resolution: they use urban population from \cite{bairoch1988} as a proxy for development, which is available for fewer cities and fewer years.

@@

The Holy Roman Empire was a powerful monarchy that existed in Central Europe from the Middle Ages until its dissolution in 1806. Though ruled by the Holy Roman Emperor at the Imperial level, it consisted of many small and large territories with their own regional rulers. Before Germany first became a republic in 1918, it is not helpful to think of "states" in the modern sense: while territories may have existed as political entities on paper, they were ruled by aristocrats who handled power in their familial structures \citep[pp. 2-3]{pt2}. These aristocratic families defined the territorial history of the HRE: territories were split by inheritances, joined by marriages and extinctions, and driven to war by struggles for dominance among the noble lineages ruling them. For this reason, when I say "state" in the context of this paper I mean its early modern German equivalent --- the territory of a lineage of rulers, or the lineage itself.

@@

I use data from the \textit{Princes and Townspeople} collection which is based on the \textit{Deutsches Städtebuch} \citep{keyser1939} and \cite{kobler2007}'s \textit{Historisches Lexikon der Deutschen Länder}. Out of the six datasets currently published in the collection, I use those on territorial histories \citep{pt2}, construction activity \citep{pt5} and conflict incidents \citep{pt6} in my analysis.\footnote{I thank Davide Cantoni and Matthias Weigand for giving me access to unpublished yearly versions of the construction and territorial histories data.} I also use the geo-referenced city locations \citep{pt1} to create visualisations.

@@

Figure \ref{fig:map_cities_raw} shows the spatial distribution of the 2,390 cities in the original \textit{Princes and Townspeople} data. The reporting boundaries are those of the German Empire in 1937 plus the Free City of Danzig; within those boundaries \cite{keyser1939} claim to list "every settlement that at one point in the history of the Holy Roman Empire was awarded the status of a city" \citep[p. 2]{pt1}.

\begin{figure}[ht]
    \centering
    \includegraphics[scale=0.15]{paper/output/descriptive/map_cities_raw.png}
    \caption{Distribution of cities in the \textit{Territorial Histories} dataset \citep{pt2}. Each dot represents a city. Some major cities are identified with labels. Polygons on the map do \textbf{not} represent historical borders; they merely represent regions by which cities are clustered in the \textit{Städtebuch}.}
    \label{fig:map_cities_raw}
\end{figure}

@@

A city-by-year panel of territorial history, taken from \cite{pt2}, forms the backbone of my dataset. For each city and each year it identifies the noble lineage to whose territory the city belonged in that year. I use this information to find years in which cities switch from one territory to another. The median city in the original dataset switches territories 5.64 times in its history.

@@

I will regress construction activity on breaks in the territorial history of cities throughout the main part of this paper --- this is not self-explanatory. After all, the research question deals with the effect of switching states on urban economic development. I use construction as a proxy for development because it is available for most cities: out of the 2,390 cities in the \textit{Städtebuch}, 2,177 have at least one construction event linked to them. By contrast, only 207 cities from the \textit{Städtebuch} appear in \cite{bairoch1988}'s dataset of city populations.

@@

Because the \textit{Städtebuch} data compiled by \cite{pt2} records exactly how each city was acquired by the territory holding it, I am able to estimate heterogeneous effects for different types of ownership changes.\footnote
{
Presumably \cite{schoenholzer2022}, having sourced their data on territorial histories from historical maps, lack detailed information about individual switching events. They instead use clever panel decomposition methods (pp. 14-20) and data on state-level characteristics (pp. 21-23) to explain what drives their estimates.
}
For this purpose I extend Equation \eqref{eq:sw22} by interacting the treatment indicator with the type of switching event:
\begin{equation}
\label{eq:baseline}
\begin{split}
    construction_{it} &= \alpha_i + \delta_t + \mathbf{X}_{it} \beta + 
    \mathbf{1}(t \ge e^{NewState}_i) \pi \\
    &+ \mathbf{1}(t \ge e^{NewState}_i) \times
    \mathbf{1}(Conquest_{it}) \cdot \theta_{conquest} \\
    &+ \mathbf{1}(t \ge e^{NewState}_i) \times
    \mathbf{1}(Other_{it}) \cdot \theta_{other} + \varepsilon_{it},
\end{split}
\end{equation}

@@

It would be a big improvement on my research strategy to find a clearer measure of state power, to check whether measurement error really does bias my estimates downward. \cite{cantoni2023} construct a novel dataset of territory-level institutions which one could potentially use for that purpose. Another improvement would be to control for the number of cities switching to another state at the same time in Equation \eqref{eq:baseline}, as described in the Discussion. Third, an event study version of Equation \eqref{eq:baseline} may be able to better tease out significant effects if the effect of state power takes some time to phase in. Finally, recent literature (\cite{baker2022} provide an overview) has questioned the validity of standard two-way fixed effects regressions with staggered adoption and time-varying treatment --- one could improve my results by using one of the robust estimators proposed by \cite{sunab2021} or \cite{cengiz2019}, for example.

>>>>

By contrast, my story is not about population dynamics but institutions (although it is compatible with Malthusianism). A longstanding tradition in economics emphasises the role of institutional factors in long-run growth (e.g. \cite{north1970}, \cite{delong1993}, \cite*{ajr2001}). From this vast literature one theme may be especially relevant here: the institution of fiscal capacity, or the state's ability to collect taxes. \cite{tilly1985} argues that the main reason for states to collect taxes is to be able to wage war. In other words, fiscal capacity and state power are tightly linked --- so if conflict selects for state power, then it also selects for fiscal capacity. \cite{gennaioli2015} and \cite*{cantoni2023} show that conflict drove state consolidation, with the latter set of authors using the same data as this paper. \cite{dincecco2012} show that fiscal capacity had a positive effect on growth.

@@

The closest relative of this paper is \cite{schoenholzer2022} who investigate how switching between states affects the development of cities. They find that switching states affects population negatively in the short run but positively in the long run. Further, they show that this long-run benefit is driven by improved state quality after switching. Similar to the evolution metaphor, they interpret this as "creative destruction": the short run cost of destruction is compensated by the long-run benefit of becoming part of a higher-quality state. However, they do not show explicitly how these improvements in state quality are driven by selection through conflict. Their data has a bigger geographical scope (all of Europe) but lower resolution: they use urban population from \cite*{bairoch1988} as a proxy for development, which is available for fewer cities and fewer years.

@@

The Holy Roman Empire was a powerful monarchy that existed in Central Europe from the Middle Ages until its dissolution in 1806. Though ruled by the Holy Roman Emperor at the Imperial level, it consisted of many small and large territories with their own regional rulers. Before Germany first became a republic in 1918, it is not helpful to think of "states" in the modern sense: while territories may have existed as political entities on paper, they were ruled by aristocrats who handled power in their familial structures \citep*[pp. 2-3]{pt2}. These aristocratic families defined the territorial history of the HRE: territories were split by inheritances, joined by marriages and extinctions, and driven to war by struggles for dominance among the noble lineages ruling them. For this reason, when I say "state" in the context of this paper I mean its early modern German equivalent --- the territory of a lineage of rulers, or the lineage itself.

@@

I use data from the \textit{Princes and Townspeople} collection which is based on the \textit{Deutsches Städtebuch} \citep{keyser1939} and \cite{kobler2007}'s \textit{Historisches Lexikon der Deutschen Länder}. Out of the six datasets currently published in the collection, I use those on territorial histories \citep*{pt2}, construction activity \citep{pt5} and conflict incidents \citep{pt6} in my analysis.\footnote{I thank Davide Cantoni and Matthias Weigand for giving me access to unpublished yearly versions of the construction and territorial histories data.} I also use the geo-referenced city locations \citep*{pt1} to create visualisations.

@@

Figure \ref{fig:map_cities_raw} shows the spatial distribution of the 2,390 cities in the original \textit{Princes and Townspeople} data. The reporting boundaries are those of the German Empire in 1937 plus the Free City of Danzig; within those boundaries \cite{keyser1939} claim to list "every settlement that at one point in the history of the Holy Roman Empire was awarded the status of a city" \citep*[p. 2]{pt1}.

\begin{figure}[ht]
    \centering
    \includegraphics[scale=0.15]{paper/output/descriptive/map_cities_raw.png}
    \caption{Distribution of cities in the \textit{Territorial Histories} dataset \citep*{pt2}. Each dot represents a city. Some major cities are identified with labels. Polygons on the map do \textbf{not} represent historical borders; they merely represent regions by which cities are clustered in the \textit{Städtebuch}.}
    \label{fig:map_cities_raw}
\end{figure}

@@

A city-by-year panel of territorial history, taken from \cite*{pt2}, forms the backbone of my dataset. For each city and each year it identifies the noble lineage to whose territory the city belonged in that year. I use this information to find years in which cities switch from one territory to another. The median city in the original dataset switches territories 5.64 times in its history.

@@

I will regress construction activity on breaks in the territorial history of cities throughout the main part of this paper --- this is not self-explanatory. After all, the research question deals with the effect of switching states on urban economic development. I use construction as a proxy for development because it is available for most cities: out of the 2,390 cities in the \textit{Städtebuch}, 2,177 have at least one construction event linked to them. By contrast, only 207 cities from the \textit{Städtebuch} appear in \cite*{bairoch1988}'s dataset of city populations.

@@

Because the \textit{Städtebuch} data compiled by \cite*{pt2} records exactly how each city was acquired by the territory holding it, I am able to estimate heterogeneous effects for different types of ownership changes.\footnote
{
Presumably \cite{schoenholzer2022}, having sourced their data on territorial histories from historical maps, lack detailed information about individual switching events. They instead use clever panel decomposition methods (pp. 14-20) and data on state-level characteristics (pp. 21-23) to explain what drives their estimates.
}
For this purpose I extend Equation \eqref{eq:sw22} by interacting the treatment indicator with the type of switching event:
\begin{equation}
\label{eq:baseline}
\begin{split}
    construction_{it} &= \alpha_i + \delta_t + \mathbf{X}_{it} \beta + 
    \mathbf{1}(t \ge e^{NewState}_i) \pi \\
    &+ \mathbf{1}(t \ge e^{NewState}_i) \times
    \mathbf{1}(Conquest_{it}) \cdot \theta_{conquest} \\
    &+ \mathbf{1}(t \ge e^{NewState}_i) \times
    \mathbf{1}(Other_{it}) \cdot \theta_{other} + \varepsilon_{it},
\end{split}
\end{equation}

@@

It would be a big improvement on my research strategy to find a clearer measure of state power, to check whether measurement error really does bias my estimates downward. \cite*{cantoni2023} construct a novel dataset of territory-level institutions which one could potentially use for that purpose. Another improvement would be to control for the number of cities switching to another state at the same time in Equation \eqref{eq:baseline}, as described in the Discussion. Third, an event study version of Equation \eqref{eq:baseline} may be able to better tease out significant effects if the effect of state power takes some time to phase in. Finally, recent literature (\cite*{baker2022} provide an overview) has questioned the validity of standard two-way fixed effects regressions with staggered adoption and time-varying treatment --- one could improve my results by using one of the robust estimators proposed by \cite{sunab2021} or \cite{cengiz2019}, for example.
